---
title: "The Three Point Revolution: NBA vs NCAA"
format: revealjs
editor: visual
---

```{r}
library(tidyverse)
library(hoopR)
library(ggplot2)

nba_raw <- hoopR::load_nba_pbp(seasons = 2025)
ncaa_raw <- hoopR::load_mbb_pbp(seasons = 2025)

# filter for only shooting plays
nba_shots <- nba_raw |> filter(shooting_play == TRUE, 
                               !str_detect(tolower(type_text), "free throw"))
ncaa_shots <- ncaa_raw |> filter(shooting_play == TRUE,
                                 !str_detect(tolower(type_text), "freethrow"))

# clean nba data
nba_clean <- nba_shots |>
  # Filter for the half-court
  filter(abs(coordinate_x) <= 47 & abs(coordinate_y) <= 25) |>
  filter(!str_detect(home_team_name, "Team")) |> # filters out all star teams
  mutate(
    league = "NBA",
    shot_outcome = scoring_play,
    team_name = if_else(team_id == home_team_id, home_team_name, away_team_name),
    pts_value = if_else(str_detect(tolower(text), "three point"), 3, 2),
    
    # Distance from the hoop at (41.75, 0)
    shot_dist = sqrt((abs(coordinate_x) - 41.75)^2 + coordinate_y^2),
    # Depth from baseline 
    court_depth = 47 - abs(coordinate_x),
    # Width from center
    court_width = coordinate_y,
    
    
    # make the court zones
    shot_zone = case_when(
      shot_dist <= 4 ~ "Rim",
      shot_dist > 4 & shot_dist <= 14 & abs(court_width) <= 8 ~ "Paint",
      
      # 3-Pointers
      pts_value == 3 & court_depth <= 14 & court_width < 0 ~ "Left Corner 3",
      pts_value == 3 & court_depth <= 14 & court_width >= 0 ~ "Right Corner 3",
      pts_value == 3 & abs(court_width) <= 8 ~ "Top of the Key 3",
      pts_value == 3 & court_width < -8 ~ "Left Wing 3",
      pts_value == 3 & court_width > 8 ~ "Right Wing 3",
      
      # Mid-Range
      pts_value == 2 & court_depth <= 14 & court_width < -8 ~ "Left Corner Mid",
      pts_value == 2 & court_depth <= 14 & court_width > 8 ~ "Right Corner Mid",
      pts_value == 2 & abs(court_width) <= 8 ~ "Mid-Range Center",
      pts_value == 2 & court_depth > 14 & court_width < -8 ~ "Mid-Range Left",
      pts_value == 2 & court_depth > 14 & court_width > 8 ~ "Right Wing Mid",
      
      TRUE ~ "Other"
    )
  ) |>
  select(athlete_id_1, league, team_name, coordinate_x, coordinate_y, shot_outcome, pts_value, shot_zone)


# lets keep only the top 100 ncaa teams
top_100 <- c(
  "Houston", "UConn", "Purdue", "Arizona", "Auburn", "Iowa State", "Tennessee", 
  "North Carolina", "Alabama", "Duke", "Creighton", "BYU", "Illinois", 
  "Marquette", "Baylor", "Saint Mary's", "Gonzaga", "Kentucky", "Kansas", 
  "San Diego State", "Wisconsin", "New Mexico", "Dayton", "Michigan State", 
  "Colorado", "Boise State", "Texas Tech", "Indiana State", "Florida", 
  "Texas", "Mississippi State", "St. John's", "Nebraska", "Nevada", "Clemson", 
  "Colorado State", "Cincinnati", "Utah State", "Florida Atlantic", "Pittsburgh", 
  "Villanova", "TCU", "Wake Forest", "Washington State", "Texas A&M", 
  "Oklahoma", "Drake", "Utah", "Ohio State", "Grand Canyon", "South Carolina", 
  "James Madison", "Northwestern", "Virginia", "Princeton", "McNeese", 
  "Bradley", "Providence", "Oregon", "Virginia Tech", "UCF", "Iowa", 
  "NC State", "Xavier", "SMU", "San Francisco", "Seton Hall", "Butler", 
  "Washington", "Kansas State", "VCU", "App State", "Richmond", "Samford", 
  "Memphis", "UNLV", "UC Irvine", "South Florida", "North Texas", "Duquesne", 
  "Boston College", "Maryland", "Yale", "Syracuse", "Loyola Chicago", 
  "Penn State", "St. Bonaventure", "USC", "Minnesota", "Ole Miss", 
  "Massachusetts", "Saint Joseph's", "George Mason", "LSU", "Cornell", 
  "Florida State", "Charleston", "Indiana", "Louisiana Tech", "Georgia"
)

# clean ncaa data
ncaa_clean <- ncaa_shots |>
  filter(abs(coordinate_x) <= 47 & abs(coordinate_y) <= 25) |>
  mutate(
    league = "NCAA",
    shot_outcome = scoring_play,
    team_name = if_else(team_id == home_team_id, home_team_name, away_team_name),
    pts_value = score_value,
    
    
    shot_dist = sqrt((abs(coordinate_x) - 41.75)^2 + coordinate_y^2),
    court_depth = 47 - abs(coordinate_x),
    court_width = coordinate_y,
    
    # zones
    shot_zone = case_when(
      shot_dist <= 4 ~ "Rim",
      shot_dist > 4 & shot_dist <= 14 & abs(court_width) <= 6 ~ "Paint",
      
      # 3-Pointers
      pts_value == 3 & court_depth <= 9.25 & court_width < 0 ~ "Left Corner 3",
      pts_value == 3 & court_depth <= 9.25 & court_width >= 0 ~ "Right Corner 3",
      pts_value == 3 & abs(court_width) <= 8 ~ "Top of the Key 3",
      pts_value == 3 & court_width < -8 ~ "Left Wing 3",
      pts_value == 3 & court_width > 8 ~ "Right Wing 3",
      
      # Mid-Range
      pts_value == 2 & court_depth <= 9.25 & court_width < -6 ~ "Left Corner Mid",
      pts_value == 2 & court_depth <= 9.25 & court_width > 6 ~ "Right Corner Mid",
      pts_value == 2 & abs(court_width) <= 8 ~ "Mid-Range Center",
      pts_value == 2 & court_depth > 9.25 & court_width < -8 ~ "Mid-Range Left",
      pts_value == 2 & court_depth > 9.25 & court_width > 8 ~ "Right Wing Mid",
      
      TRUE ~ "Other"
    )
  ) |>
  filter(team_name %in% top_100) |>
  select(athlete_id_1, league, team_name, coordinate_x, coordinate_y, court_width, court_depth, shot_outcome, pts_value, shot_zone)

# combine nba and ncaa into 1 df 
shot_df <- bind_rows(nba_clean, ncaa_clean)

# look at some stats per zone
zone_efficiency <- shot_df |>
  group_by(league, shot_zone) |>
  summarise(
    shots_taken = n(),
    fg_pct = mean(shot_outcome),
    points_per_shot = mean(shot_outcome * pts_value)
  ) |>
  arrange(shot_zone, league) |>
  mutate(shot_zone = factor(shot_zone, levels = c("Rim", "Paint", "Left Corner Mid", "Right Corner Mid", "Mid-Range Left", "Right Wing Mid", "Mid-Range Center", "Left Corner 3", "Right Corner 3", "Left Wing 3", "Right Wing 3", "Top of the Key 3")))
```


## Data Introduction

- play by play data from 2025 season for both NBA and NCAA
- data is from hoopR R package
- categorizing shots into 12 zones on the court

## Questions of Interest

- Has the three point revolution effected NBA or NCAA teams more?
- Is the approach of shooting more threes correct? 
- Does it generate more efficient offense?

## Are Teams Shooting More Threes?

```{r}
# Load 10 years of Team Box Scores - because it's smaller than pbp data 
seasons <- 2016:2025

nba_box <- hoopR::load_nba_team_box(seasons = seasons)
ncaa_box <- hoopR::load_mbb_team_box(seasons = seasons)


# Calculate NBA Trends
nba_trend <- nba_box |>
  filter(field_goals_attempted > 0) |> 
  mutate(three_p_ar = three_point_field_goals_attempted / field_goals_attempted) |>
  group_by(season) |>
  summarize(three_rate = mean(three_p_ar, na.rm = TRUE), league = "NBA")

# Calculate NCAA Trends
ncaa_trend <- ncaa_box |>
  filter(field_goals_attempted > 0) |>
  mutate(three_p_ar = three_point_field_goals_attempted / field_goals_attempted) |>
  group_by(season) |>
  summarize(three_rate = mean(three_p_ar, na.rm = TRUE), league = "NCAA")



# Make the Plot
bind_rows(nba_trend, ncaa_trend) |>
  ggplot(aes(x = season, y = three_rate, color = league, group = league)) +
  geom_line(linewidth = 1.2) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Are teams Shooting More threes?",
    y = "3-Point Attempt Rate (% of Total Shots)",
    x = "Season",
    color = "League"
  ) +
  scale_color_viridis_d() +
  theme_minimal(base_size = 18) 
```

## Shot Distribution

```{r}
# Compute shot count per zone per league
zone_counts <- shot_df |>
  count(league, shot_zone) |>
  group_by(league) |>
  mutate(shot_pct = n / sum(n) * 100)  # % of shots from that zone

# Join back so every point knows its zone's shot %
shot_df <- shot_df |>
  left_join(zone_counts, by = c("league", "shot_zone"))

# Plot color by shot_pct, increase point size to fill zones visually
ggplot(shot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = shot_pct)) +
  geom_point(size = 1.1, alpha = 0.8) +
  facet_wrap(~league) +
  scale_color_gradient(
    name = "% of Shots",
    low = "#fde8e8",
    high = "#8b0000"
  ) + 
  coord_fixed() +
  theme_minimal(base_size = 24) +
  labs(
    title = "Shot Distribution by Zone",
    x = NULL, y = NULL
  )
```

## Efficiency 

```{r}
zone_pts <- shot_df |>
  group_by(league, shot_zone) |>
  summarise(pts_per_shot = mean(shot_outcome * pts_value, na.rm = TRUE), .groups = "drop")

shot_df <- shot_df |>
  left_join(zone_pts, by = c("league", "shot_zone"))

ggplot(shot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = pts_per_shot)) +
  geom_point(size = 1.1, alpha = 0.8) +
  facet_wrap(~league) +
  scale_color_gradient(
    name = "Pts/Shot",
    low = "#fde8e8",
    high = "#8b0000"
  ) +
  coord_fixed() +
  theme_minimal(base_size = 24) +
  labs(
    title = "Points per Shot by Zone",
    x = NULL, y = NULL
  )
```

