---
title: "index"
format: html
---

# Introduction
Over the past decade, basketball has been going through a shift, largely due to analytics. Teams have been viewing the three point shot as more and more valuable, while the mid-range slowly becomes a lost art. Throughout basketball history, people have always been encouraged to take a step in if they have a better chance of making the shot. However, as of late the game has shifted into the idea of getting as many threes and layups as possible.

Public opinion seems to say that this change has mainly effected the NBA, while college basketball sticks to traditional, fundamental game plans. People seem to think college teams, while obviously shooting more threes than they used to, haven't fully subscribed to this new ideology of abandoning the mid range. 

For this project, I'm using play-by-play data from the 2024-2025 NBA and NCAA seasons(found in the hoopR R package) to examine shot selection and efficiency at both levels of basketball. I want to determine if this evolution of the three pointer has really effected the NBA more than the college level, and also if this shift in ideology is even the right approach to begin with. Should teams continue to try and maximize their efficiency by eliminating mid range shots, or are teams that subscribe to this philosophy hurting themselves?

To tackle these questions, I'm going to be breaking the court into 12 hotzones, 5 zones behind the three point line, 5 zones in the midrange, a zone for shots at the rim, and a zone for other shots in the paint. I'm going to be looking at all 30 NBA teams from the 2025 season, but I'm filtering out some college teams so we are only looking at the top 100 teams from the 2025 season. That way we can filter out any weird outliers or really bad teams. 


# first lets check if teams are shooting more threes
```{r}

# 1. Load 10 years of Team Box Scores (much smaller than PBP)
seasons <- 2016:2025

# For NBA
nba_box <- hoopR::load_nba_team_box(seasons = seasons)
ncaa_box <- hoopR::load_mbb_team_box(seasons = seasons)


# 1. Calculate NBA Trends
nba_trend <- nba_box |>
  filter(field_goals_attempted > 0) |> 
  mutate(three_p_ar = three_point_field_goals_attempted / field_goals_attempted) |>
  group_by(season) |>
  summarize(three_rate = mean(three_p_ar, na.rm = TRUE), league = "NBA")

# 2. Calculate NCAA Trends (Assuming ncaa_box has the same names)
ncaa_trend <- ncaa_box |>
  filter(field_goals_attempted > 0) |>
  mutate(three_p_ar = three_point_field_goals_attempted / field_goals_attempted) |>
  group_by(season) |>
  summarize(three_rate = mean(three_p_ar, na.rm = TRUE), league = "NCAA")



# 3. Combine and Plot
bind_rows(nba_trend, ncaa_trend) |>
  ggplot(aes(x = season, y = three_rate, color = league, group = league)) +
  geom_line(linewidth = 1.2) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "The 10-Year 3-Point Revolution",
    subtitle = "Rise of 3-Point Attempt Rate (3PAr) in Pro vs. Elite College Ball",
    y = "3-Point Attempt Rate (% of Total Shots)",
    x = "Season",
    color = "League"
  ) +
  scale_color_viridis_d() +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "The 10-Year 3-Point Revolution",
    subtitle = "Rise of 3-Point Attempt Rate (3PAr) in Pro vs. Elite College Ball",
    y = "3-Point Attempt Rate (% of Total Shots)",
    x = "Season",
    color = "League"
  )
```




```{r}
library(tidyverse)
library(hoopR)
library(ggplot2)

nba_raw <- hoopR::load_nba_pbp(seasons = 2025)
ncaa_raw <- hoopR::load_mbb_pbp(seasons = 2025)

# filter for only shooting plays
nba_shots <- nba_raw |> filter(shooting_play == TRUE, 
                               !str_detect(tolower(type_text), "free throw"))
ncaa_shots <- ncaa_raw |> filter(shooting_play == TRUE,
                                 !str_detect(tolower(type_text), "freethrow"))

# clean nba data
nba_clean <- nba_shots |>
  # Filter for the half-court
  filter(abs(coordinate_x) <= 47 & abs(coordinate_y) <= 25) |>
  filter(!str_detect(home_team_name, "Team")) |> # filters out all star teams
  mutate(
    league = "NBA",
    shot_outcome = scoring_play,
    team_name = if_else(team_id == home_team_id, home_team_name, away_team_name),
    pts_value = if_else(str_detect(tolower(text), "three point"), 3, 2),
    
    # Distance from the hoop at (41.75, 0)
    shot_dist = sqrt((abs(coordinate_x) - 41.75)^2 + coordinate_y^2),
    # Depth from baseline 
    court_depth = 47 - abs(coordinate_x),
    # Width from center
    court_width = coordinate_y,
    
    shot_zone = case_when(
      shot_dist <= 4 ~ "Rim",
      shot_dist > 4 & shot_dist <= 14 & abs(court_width) <= 8 ~ "Paint",
      
      # 3-Pointers
      pts_value == 3 & court_depth <= 14 & court_width < 0 ~ "Left Corner 3",
      pts_value == 3 & court_depth <= 14 & court_width >= 0 ~ "Right Corner 3",
      pts_value == 3 & abs(court_width) <= 8 ~ "Top of the Key 3",
      pts_value == 3 & court_width < -8 ~ "Left Wing 3",
      pts_value == 3 & court_width > 8 ~ "Right Wing 3",
      
      # Mid-Range
      pts_value == 2 & court_depth <= 14 & court_width < -8 ~ "Left Corner Mid",
      pts_value == 2 & court_depth <= 14 & court_width > 8 ~ "Right Corner Mid",
      pts_value == 2 & abs(court_width) <= 8 ~ "Mid-Range Center",
      pts_value == 2 & court_depth > 14 & court_width < -8 ~ "Mid-Range Left",
      pts_value == 2 & court_depth > 14 & court_width > 8 ~ "Right Wing Mid",
      
      TRUE ~ "Other"
    )
  ) |>
  select(athlete_id_1, league, team_name, coordinate_x, coordinate_y, shot_outcome, pts_value, shot_zone)


# lets keep only the top 100 ncaa teams
top_100 <- c(
  "Houston", "UConn", "Purdue", "Arizona", "Auburn", "Iowa State", "Tennessee", 
  "North Carolina", "Alabama", "Duke", "Creighton", "BYU", "Illinois", 
  "Marquette", "Baylor", "Saint Mary's", "Gonzaga", "Kentucky", "Kansas", 
  "San Diego State", "Wisconsin", "New Mexico", "Dayton", "Michigan State", 
  "Colorado", "Boise State", "Texas Tech", "Indiana State", "Florida", 
  "Texas", "Mississippi State", "St. John's", "Nebraska", "Nevada", "Clemson", 
  "Colorado State", "Cincinnati", "Utah State", "Florida Atlantic", "Pittsburgh", 
  "Villanova", "TCU", "Wake Forest", "Washington State", "Texas A&M", 
  "Oklahoma", "Drake", "Utah", "Ohio State", "Grand Canyon", "South Carolina", 
  "James Madison", "Northwestern", "Virginia", "Princeton", "McNeese", 
  "Bradley", "Providence", "Oregon", "Virginia Tech", "UCF", "Iowa", 
  "NC State", "Xavier", "SMU", "San Francisco", "Seton Hall", "Butler", 
  "Washington", "Kansas State", "VCU", "App State", "Richmond", "Samford", 
  "Memphis", "UNLV", "UC Irvine", "South Florida", "North Texas", "Duquesne", 
  "Boston College", "Maryland", "Yale", "Syracuse", "Loyola Chicago", 
  "Penn State", "St. Bonaventure", "USC", "Minnesota", "Ole Miss", 
  "Massachusetts", "Saint Joseph's", "George Mason", "LSU", "Cornell", 
  "Florida State", "Charleston", "Indiana", "Louisiana Tech", "Georgia"
)

# clean ncaa data
ncaa_clean <- ncaa_shots |>
  filter(abs(coordinate_x) <= 47 & abs(coordinate_y) <= 25) |>
  mutate(
    league = "NCAA",
    shot_outcome = scoring_play,
    team_name = if_else(team_id == home_team_id, home_team_name, away_team_name),
    pts_value = score_value,
    
    
    shot_dist = sqrt((abs(coordinate_x) - 41.75)^2 + coordinate_y^2),
    court_depth = 47 - abs(coordinate_x),
    court_width = coordinate_y,
    
    # zones
    shot_zone = case_when(
      shot_dist <= 4 ~ "Rim",
      shot_dist > 4 & shot_dist <= 14 & abs(court_width) <= 6 ~ "Paint",
      
      # 3-Pointers
      pts_value == 3 & court_depth <= 9.25 & court_width < 0 ~ "Left Corner 3",
      pts_value == 3 & court_depth <= 9.25 & court_width >= 0 ~ "Right Corner 3",
      pts_value == 3 & abs(court_width) <= 8 ~ "Top of the Key 3",
      pts_value == 3 & court_width < -8 ~ "Left Wing 3",
      pts_value == 3 & court_width > 8 ~ "Right Wing 3",
      
      # Mid-Range
      pts_value == 2 & court_depth <= 9.25 & court_width < -6 ~ "Left Corner Mid",
      pts_value == 2 & court_depth <= 9.25 & court_width > 6 ~ "Right Corner Mid",
      pts_value == 2 & abs(court_width) <= 8 ~ "Mid-Range Center",
      pts_value == 2 & court_depth > 9.25 & court_width < -8 ~ "Mid-Range Left",
      pts_value == 2 & court_depth > 9.25 & court_width > 8 ~ "Right Wing Mid",
      
      TRUE ~ "Other"
    )
  ) |>
  filter(team_name %in% top_100) |>
  select(athlete_id_1, league, team_name, coordinate_x, coordinate_y, court_width, court_depth, shot_outcome, pts_value, shot_zone)

# combine nba and ncaa into 1 df 
shot_df <- bind_rows(nba_clean, ncaa_clean)

# look at some stats per zone
zone_efficiency <- shot_df |>
  group_by(league, shot_zone) |>
  summarise(
    shots_taken = n(),
    fg_pct = mean(shot_outcome),
    points_per_shot = mean(shot_outcome * pts_value)
  ) |>
  arrange(shot_zone, league) |>
  mutate(shot_zone = factor(shot_zone, levels = c("Rim", "Paint", "Left Corner Mid", "Right Corner Mid", "Mid-Range Left", "Right Wing Mid", "Mid-Range Center", "Left Corner 3", "Right Corner 3", "Left Wing 3", "Right Wing 3", "Top of the Key 3")))

# try making a plot
ggplot(shot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = shot_zone)) +
  geom_point(alpha = 0.1, size = 0.5) +
  facet_wrap(~league) +
  theme_minimal() +
  coord_fixed() +
  scale_color_viridis_d() 


# make a bar plot comparing points per zone
ggplot(zone_efficiency, aes(x = shot_zone, y = points_per_shot, fill = league)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_viridis_d() +
  theme_minimal()


```


# shots per zone
```{r}
# Compute shot count per zone per league
zone_counts <- shot_df |>
  count(league, shot_zone) |>
  group_by(league) |>
  mutate(shot_pct = n / sum(n) * 100)  # % of shots from that zone

# Join back so every point knows its zone's shot %
shot_df <- shot_df |>
  left_join(zone_counts, by = c("league", "shot_zone"))

# Plot color by shot_pct, increase point size to fill zones visually
ggplot(shot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = shot_pct)) +
  geom_point(size = 1.1, alpha = 0.8) +
  facet_wrap(~league) +
  scale_color_gradient(
    name   = "% of Shots",
    low    = "white",
    high   = "#8b0000",
    labels = scales::label_number(suffix = "%", accuracy = 0.1)
  ) + 
  coord_fixed() +
  theme_minimal() +
  labs(
    title = "Shot Distribution by Zone: NBA vs. NCAA",
    x = NULL, y = NULL
  )
```

#fg%
```{r}
zone_fg <- shot_df |>
  group_by(league, shot_zone) |>
  summarise(fg_pct = mean(shot_outcome, na.rm = TRUE) * 100, .groups = "drop")

shot_df <- shot_df |>
  left_join(zone_fg, by = c("league", "shot_zone"))

ggplot(shot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = fg_pct)) +
  geom_point(size = 1.1, alpha = 0.8) +
  facet_wrap(~league) +
  scale_color_gradient(
    name   = "FG%",
    low    = "white",
    high   = "#8b0000",
    labels = scales::label_number(suffix = "%", accuracy = 0.1)
  ) +
  coord_fixed() +
  theme_minimal() +
  labs(
    title = "Field Goal % by Zone: NBA vs. NCAA",
    x = NULL, y = NULL
  )
```


# pts per shot
```{r}
zone_pts <- shot_df |>
  group_by(league, shot_zone) |>
  summarise(pts_per_shot = mean(shot_outcome * pts_value, na.rm = TRUE), .groups = "drop")

shot_df <- shot_df |>
  left_join(zone_pts, by = c("league", "shot_zone"))

ggplot(shot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = pts_per_shot)) +
  geom_point(size = 1.1, alpha = 0.8) +
  facet_wrap(~league) +
  scale_color_gradient(
    name   = "Pts/Shot",
    low    = "white",
    high   = "#8b0000",
    labels = scales::label_number(accuracy = 0.01)
  ) +
  coord_fixed() +
  theme_minimal() +
  labs(
    title = "Points per Shot by Zone: NBA vs. NCAA",
    x = NULL, y = NULL
  )
```

# efficiency
```{r}
zone_pts_wide <- shot_df |>
  group_by(league, shot_zone) |>
  summarise(pts_per_shot = mean(shot_outcome * pts_value, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = league, values_from = pts_per_shot) |>
  mutate(shot_zone = fct_reorder(shot_zone, NBA))  # sort by NBA value

ggplot(zone_pts_wide, aes(y = shot_zone)) +
  geom_segment(aes(x = NCAA, xend = NBA, yend = shot_zone), color = "#cccccc", linewidth = 1) +
  geom_point(aes(x = NBA),  color = "#8b0000", size = 4) +
  geom_point(aes(x = NCAA), color = "#d4a0a0", size = 4) +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "#999999", linewidth = 0.7) +
  annotate("text", x = 1.01, y = 0.6, label = "1.0 pts/shot", 
           hjust = 0, size = 3, color = "#999999") +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.01)) +
  theme_minimal() +
  labs(
    title    = "Efficiency by Zone: NBA vs. NCAA",
    subtitle = "Dark red = NBA · Pink = NCAA · Dashed line = 1.0 pts/shot break-even",
    x = "Points per Shot", y = NULL
  )
```





# get players for shiny app
```{r}
nba_rosters  <- hoopR::load_nba_player_box(seasons = 2025) |>
  select(athlete_id, athlete_display_name) |> distinct()

ncaa_rosters <- hoopR::load_mbb_player_box(seasons = 2025) |>
  select(athlete_id, athlete_display_name) |> distinct()

rosters <- bind_rows(nba_rosters, ncaa_rosters) |> distinct()

shot_df <- shot_df |>
  left_join(rosters, by = c("athlete_id_1" = "athlete_id"))
```



# shiny app
```{r}
library(shiny)


ui <- fluidPage(
  titlePanel("NBA vs. NCAA Shot Zone Explorer"),
  sidebarLayout(
    sidebarPanel(
      radioButtons("compare_type", "Compare by:",
                   choices = c("Teams", "Players")),
      selectInput("metric", "Metric:",
                  choices = c("% of Shots" = "shot_pct",
                              "FG%"         = "fg_pct",
                              "Pts/Shot"    = "pts_per_shot")),
      hr(),
      strong("Selection 1"),
      radioButtons("league1", "League:", choices = c("NBA", "NCAA")),
      selectizeInput("selection1", "Team/Player:", choices = NULL),
      hr(),
      strong("Selection 2"),
      radioButtons("league2", "League:", choices = c("NBA", "NCAA")),
      selectizeInput("selection2", "Team/Player:", choices = NULL),
    ),

    mainPanel(
      fluidRow(
        column(6, plotOutput("plot1", height = "500px")),
        column(6, plotOutput("plot2", height = "500px"))
      )
    )
  )
)

server <- function(input, output, session) {
  
  # when league1 or compare_type changes, update selection1 dropdown
  observeEvent(c(input$league1, input$compare_type), {
    df <- shot_df |> filter(league == input$league1)
    if (input$compare_type == "Teams") {
      new_choices <- df |> distinct(team_name) |> pull(team_name) |> sort()
    } else {
      new_choices <- df |> filter(!is.na(athlete_display_name)) |>
        distinct(athlete_display_name) |> pull(athlete_display_name) |> sort()
    }
    updateSelectizeInput(inputId = "selection1", choices = new_choices)
  })

  # same for selection2
  observeEvent(c(input$league2, input$compare_type), {
    df <- shot_df |> filter(league == input$league2)
    if (input$compare_type == "Teams") {
      new_choices <- df |> distinct(team_name) |> pull(team_name) |> sort()
    } else {
      new_choices <- df |> filter(!is.na(athlete_display_name)) |>
        distinct(athlete_display_name) |> pull(athlete_display_name) |> sort()
    }
    updateSelectizeInput(inputId = "selection2", choices = new_choices)
  })
  
  # reactive 1
  reactive_df1 <- reactive({
    df <- shot_df |> filter(league == input$league1)
    if (input$compare_type == "Teams") {
      df |> filter(team_name == input$selection1)
    } else {
      df |> filter(athlete_display_name == input$selection1)
    }
  })

  # reactive 2
  reactive_df2 <- reactive({
    df <- shot_df |> filter(league == input$league2)
    if (input$compare_type == "Teams") {
      df |> filter(team_name == input$selection2)
    } else {
      df |> filter(athlete_display_name == input$selection2)
    }
  })
  
  
  # plot 1
  output$plot1 <- renderPlot({
    
    # labels for plot titles
    metric_label <- case_when(
      input$metric == "shot_pct" ~ "% of Total Shots",
      input$metric == "fg_pct" ~ "FG%",
      input$metric == "pts_per_shot" ~ "Expected Points Per Shot"
    )
  
    zone_stats <- reactive_df1() |>
      group_by(shot_zone) |>
      summarise(
        n = n(),
        low_sample  = n() < 15,
        fg_pct = mean(shot_outcome, na.rm = TRUE) * 100,
        pts_per_shot = mean(shot_outcome * pts_value, na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(
        shot_pct = n / sum(n) * 100,
        value = if (input$metric == "shot_pct") shot_pct
                else if (input$metric == "fg_pct") fg_pct
                else pts_per_shot
      )
  
    plot_df <- reactive_df1() |> left_join(zone_stats, by = "shot_zone")
  
    ggplot(plot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = value)) +
      geom_point(aes(alpha = if_else(low_sample, 0.15, 0.8)), size = 1.1) +
      scale_color_gradient(name = input$metric, low = "#fde8e8", high = "#8b0000") +
      scale_alpha_identity() +
      coord_fixed() +
      theme_minimal() +
      labs(title = glue::glue(input$selection1, " ", metric_label), 
           x = NULL, 
           y = NULL,
           caption = "Zones with less than 15 shots will be faded.")
  })

  # plot 2
  output$plot2 <- renderPlot({
    
    # labels for plot titles
    metric_label <- case_when(
      input$metric == "shot_pct" ~ "% of Total Shots",
      input$metric == "fg_pct" ~ "FG%",
      input$metric == "pts_per_shot" ~ "Expected Points per Shot"
    )
  
    zone_stats <- reactive_df2() |>
      group_by(shot_zone) |>
      summarise(
        n = n(),
        low_sample = n() < 15,
        fg_pct = mean(shot_outcome, na.rm = TRUE) * 100,
        pts_per_shot = mean(shot_outcome * pts_value, na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(
        shot_pct = n / sum(n) * 100,
        value = if (input$metric == "shot_pct") shot_pct
                else if (input$metric == "fg_pct") fg_pct
                else pts_per_shot
      )
  
  plot_df <- reactive_df2() |> left_join(zone_stats, by = "shot_zone")
  
  ggplot(plot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = value)) +
    geom_point(aes(alpha = if_else(low_sample, 0.15, 0.8)), size = 1.1) +
    scale_color_gradient(name = input$metric, low = "#fde8e8", high = "#8b0000") +
    scale_alpha_identity() +
    coord_fixed() +
    theme_minimal() +
    labs(title = glue::glue(input$selection2, " ", metric_label), 
         x = NULL, 
         y = NULL,
         caption = "Zones with less than 15 shots will be faded.")
  })
  
}

shinyApp(ui, server)
```


```{r}
# code for a plot
test_df <- shot_df |> filter(team_name == "Boston", league == "NBA")

total_shots <- nrow(test_df)

zone_stats <- test_df |>
  group_by(shot_zone) |>
  summarise(
    n          = n(),
    value      = n() / total_shots * 100,
    low_sample = n() < 15,
    .groups    = "drop"
  )

plot_df <- test_df |> left_join(zone_stats, by = "shot_zone")

ggplot(plot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = value)) +
  geom_point(aes(alpha = if_else(low_sample, 0.15, 0.8)), size = 1.1) +
  scale_color_gradient(
    name = "% of Shots",
    low  = "white", high = "#8b0000"
  ) +
  scale_alpha_identity() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Boston", x = NULL, y = NULL)
```

```{r}
# test fg% for boston
zone_stats <- test_df |>
  group_by(shot_zone) |>
  summarise(
    n          = n(),
    low_sample = n() < 15,
    value      = mean(shot_outcome, na.rm = TRUE) * 100,
    .groups    = "drop"
  )

plot_df <- test_df |> left_join(zone_stats, by = "shot_zone")

ggplot(plot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = value)) +
  geom_point(aes(alpha = if_else(low_sample, 0.15, 0.8)), size = 1.1) +
  scale_color_gradient(name = "FG%", low = "#fde8e8", high = "#8b0000") +
  scale_alpha_identity() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Boston", x = NULL, y = NULL)
```

```{r}
# test pts per shot
zone_stats <- test_df |>
  group_by(shot_zone) |>
  summarise(
    n          = n(),
    low_sample = n() < 15,
    value      = mean(shot_outcome * pts_value, na.rm = TRUE),
    .groups    = "drop"
  )

plot_df <- test_df |> left_join(zone_stats, by = "shot_zone")

ggplot(plot_df, aes(x = coordinate_y, y = 47 - abs(coordinate_x), color = value)) +
  geom_point(aes(alpha = if_else(low_sample, 0.15, 0.8)), size = 1.1) +
  scale_color_gradient(name = "Pts/Shot", low = "#fde8e8", high = "#8b0000") +
  scale_alpha_identity() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "Boston", x = NULL, y = NULL)
```

